---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# spline-pr

The goal of spline-pr is to fit poisson regression models with generalized estimating equations (i.e., robust variance estimators for prevalence ratios) that include spline terms. The code in this repo also generates plots that show the estimated spline curve over a domain specified by the user.

# Example

This example uses NHANES data to create a spline figure showing the prevalence ratio for blood pressure control as a function of age. Since we are dealing with blood pressure control, we filter the data to include only participants with self-reported antihypertensive medication use. Also, since this is not a 'survey' tutorial, we do not engage with the NHANES survey weights.

```{r, message=FALSE}

library(table.glue)
library(tidyverse)
library(splines)
library(geepack)
library(mice)

source("R/functions.R")

data_nhanes <- table.glue::nhanes %>% 
 as_tibble() %>% 
 select(age, bp_sys_mmhg, bp_dia_mmhg, meds_bp) %>% 
 filter(meds_bp == 'yes', age >= 20)

data_nhanes

```

## Imputation of blood pressure values

As with almost all data, we have missing values. Thus, we will use multiple imputation and pool our spline estimates from each imputed dataset.

```{r}

n_imputes <- 5

data_nhanes_impute <- data_nhanes %>% 
 # remove constant column (all meds_bp == 'yes')
 select(-meds_bp) %>% 
 # run the imputation models
 mice(method = 'pmm', m = n_imputes, printFlag = FALSE) %>% 
 # get the imputed data, action = 'all' means return every imputed data set
 complete(action = 'all') %>% 
 # convert each imputed dataset to a tibble for better printing
 map(as_tibble)

# print the first imputed dataset 
data_nhanes_impute[[1]]

```

## Post-processing of imputed values

Let's not forget that our outcome is dependent on the two blood pressure variables; i.e., blood pressure control is 1 if systolic blood pressure is < 140 mm Hg and diastolic blood pressure is < 90 mm Hg. .

```{r}

data_nhanes_impute <- data_nhanes_impute %>% 
 map(
  ~ .x %>% 
   mutate(
    bp_control = if_else(
     condition = bp_sys_mmhg < 140 & bp_dia_mmhg < 90,
     true = 1,
     false = 0
    )
   )
 )

data_nhanes_impute[[1]]

```


## Model fitting

Next we fit a `geeglm` model to each imputed dataset. Note that 

```{r}

fits <- map(
  .x = data_nhanes_impute,
  .f = ~ geeglm(bp_control ~ ns(age, df = 4), 
                data = .x, 
                id = seq(nrow(.x)))
)

summary(fits[[1]])

```

## Spline pooling

```{r}

# determine a spline basis function
# the terms used here must match those in geeglm()
bases <- map(
  .x = data_nhanes_impute,
  .f = ~ ns(.x$age, df = 4)
)

# using the fit and basis function,
# get a predicted spline estimate + SE
spline_preds <- map2(
  .x = fits, 
  .y = bases, 
  .f = get_spline_preds,
  pattern = '^ns\\(',
  x_min = 20, 
  x_max = 80,
  x_ref = 50
)

# pool results using Rubin's rules

# V_w = mean of the variance estimates
variance_within <- map_dfc(spline_preds, "se") %>% 
  apply(MARGIN = 1, function(x) mean(x^2))

# V_b = variance of the predictions
variance_between <- map_dfc(spline_preds, "pred") %>% 
  apply(MARGIN = 1, var)

# V_total = V_w + V_b * (n_imputes+1) / n_imputes
variance_total <- 
  variance_within + variance_between + variance_between/n_imputes

se_pooled <- sqrt(variance_total)

spline_pool <- tibble(
  x = spline_preds[[1]]$x,
  pred = apply(map_dfc(spline_preds, "pred"), 1, mean),
  se = se_pooled,
  ci_lwr = pred + qnorm(0.025) * se,
  ci_upr = pred + qnorm(0.975) * se
)

spline_pool

```

## Spline visualization

```{r, dpi=300, fig.width=7, fig.height=5}

data_imputed_stack <- bind_rows(data_nhanes_impute)

data_segment <- bin_segments(x = data_imputed_stack$age,
                             y = data_imputed_stack$bp_control,
                             x_min = 20,
                             x_max = 80,
                             by_y = TRUE,
                             bin_length = 1,
                             bin_count = 60,
                             bin_yintercept = 2/3) %>%
  mutate(event_status = factor(event_status,
                               levels = c(1, 0),
                               labels = c("Yes", "No")))


fig <- ggplot(spline_pool) + 
 aes(x = x, 
     y = exp(pred), 
     ymin = exp(ci_lwr), 
     ymax = exp(ci_upr)) + 
 labs(x = 'age, years',
      y = 'Prevalence ratio',
      color = 'Blood pressure control') + 
 geom_line() + 
 geom_ribbon(alpha = 0.2) + 
 scale_y_log10() + 
 geom_segment(data = data_segment, 
              inherit.aes = FALSE,
              size = 2,
              mapping = aes(x = x, 
                            y = y,
                            color = event_status,
                            xend = xend, 
                            yend = yend)) + 
 theme_bw() + 
 geom_hline(yintercept = 1, linetype = 2, color = 'grey') + 
 theme(panel.grid = element_blank(),
       legend.position = c(.2, 0.42)) + 
 scale_color_manual(values = c("grey", "black"))

fig

```






